\chapter*{Acknowledgements}

\textbfit{To my supervisors, Lars and Michael.} You guys are rock stars! Getting to hang out with you these past years has been truly wonderful, and I’m grateful for all the laughs, digressions and light-hearted discussions. Thank you for always being so patient, kind and welcoming. It's been a priviledge, and I’m really gonna miss it. \\

When it comes to the technical side, your practical and theoretical knowledge has been invaluable. Thank you, again, for lending me so much of your time, for all the feedback, and for helping me out whenever I got sidetracked. \\

\textbfit{To all my friends at IFI.} Thank you for making these five years so enjoyable. This paragraph was never gonna do you justice, but you know who you are, and I feel very fortunate. I will always remember the memories we made, and I can’t wait to make new ones. A special mention to my friends Birk Hogstad and Tarun Sharma. Without you guys, I probably wouldn't even have passed my first semester! \\

An extra shoutout to all the student organisations that make IFI so special. In particular, MAPS (for having the best events and tastiest pizza), CYB (for letting me pour beers and serve coffees), and FIFI (for hosting the annual football tournament that we kept winning.. Go Beavers!). \\

\textbfit{To the wonderful people I encountered in Italy.} For the past five months I’ve been travelling around this beautiful country, whilst working on my thesis. Thank you for all the adventures, the good conversations, and all the wine. I'm looking forward to the day we meet again! \\

Also a huge thank you to Magne Davidsen, for letting me borrow this computer! \\

\textbfit{To my family.} Thank you for your constant encouragement and patience. A big shoutout to my cousin Oljana, for landing her first job in economics after many years of being a nurse. Your constant hard work is such an inspiration to me, and I’m so proud of you! Thank you, Katja and Emil, for not hindering my attempts to make Fillip and Jakob Chelsea fans. I haven’t given up yet! \\

\hfill \textit{Sergey}